{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-11 thesis/tm.tex
tm: abort codes / retry sets <--> estimation; also some more 10hrs

Each preemption point ({\tt struct hax} as defined in {\tt tree.h})
includes three new fields.
The boolean {\tt h->xbegin} is set if the preemption point occurred at an {\tt \_xbegin()} call.
Then if set (as the tag of an option type),
the twin lists of integers
{\tt h->xabort\_codes\_ever}
and
{\tt h->xabort\_codes\_todo}
store possible error codes this {\tt \_xbegin()} call should be tested to possibly return.
The former list stores all such error codes, whether already tested or yet to be tested,
while the latter serves as a workqueue that indicates only those not already tested yet
(serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).

@@

To address this,
I extended Landslide
with {\em retry set reduction} (commit 86657c7),
an experimental feature named after prior work's analogous {\em sleep sets}
(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
if either or both sides of a memory conflict were part of a transaction,
it records a {\em retry set},
i.e., the pair of {\tt xbegin} results executed by the conflicting threads.
% ah-- not compatible after all!
%in addition to tracking the set of abort codes possible at each preemption point.
Later, when traversing that new branch,
Landslide remembers those threads' expected {\tt xbegin} results
and refuses to test any others
(unless DPOR separately found them, too, to conflict),
thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
%The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
Upcoming in the evaluation, Figure~\ref{fig:retry-reduction} will visualize the reduction achieved in two test cases.

@@

@@

\begin{table}[t]
	\begin{center}
		\footnotesize
		\begin{tabular}{cc}
		\begin{tabular}{cc||r|r}
			& & \multicolumn{2}{c}{\bf STM ({\tt -A -S})} \\
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.41} & \ints{4} \\
			({\tt spinlock})
			& 2,2 & \cpu{198.77} & \ints{1702} \\
			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
			& 3,1 & \cpu{35.28} & \ints{246} \\
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.55} & \ints{4} \\
			({\tt spin\_fixed})
			& 2,2 & \cpu{105.57} & \ints{998} \\
			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 3,1 & \cpu{28.27} & \ints{186} \\
			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \cpu{qqq} & \ints {qqq} \\ % was: 30408, 282084 pre mkrun
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.44} & \ints{4}		\\
			({\tt mutex})
			& 2,2 & \cpu{16.83} & \ints{180}	\\
			& 2,3 & \cpu{405.21} & \ints{9372}	\\
			& 2,4 & \cpu{qqq} & \ints{qqq}	\\ % was: 15266 sec 384512 pre mkrun
			& 3,1 & \cpu{15.00} & \ints{132}	\\
			%& 3,2 & \cpu{29432.87} & \ints{707086}	\\ % pre mkrun fix (let alone bottom1)
			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \cpu{665.89} & \ints{15064}	\\
			\hline
			{\tt lock\_fast}
			& 2,1 & \cpu{3.25} & \ints{1}	\\
			({\tt spin\_fixed})
			& 9,9 & \cpu{4.61} & \ints{1}	\\
			\hline
			{\tt lock\_fast}
			& 2,1 & \cpu{3.19} & \ints{1}	\\
			({\tt mutex})
			& 9,9 & \cpu{4.62} & \ints{1}	\\
			\end{tabular}
			&
			\begin{tabular}{cc||r|r}
			& & \multicolumn{2}{c}{\bf Non-transactional} \\
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt avl\_mutex}
			& 2,1 & \cpu{3.48} & \ints{7} \\
			& 2,2 & \cpu{6.25} & \ints{85} \\
			& 2,3 & \cpu{24.69} & \ints{561} \\
			& 2,4 & \cpu{217.98} & \ints{4984} \\
			& 3,1 & \cpu{8.26} & \ints{129} \\
			& 3,2 & \cpu{1403.46} & \ints{30653} \\
			& 3,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \cpu{199.96} & \ints{4488} \\
			& 4,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			\cline{1-4}
			{\tt map\_mutex}
			& 2,1 & \cpu{39.81} & \ints{83} \\
			& 2,2 & \ETAdag{26h 21m} & \ETAdag{1085126} \\
			% [JOB 9] progress: 496530/1085126 brs (45.757812%), ETA 16h 21m 0s (elapsed 9h 59m 38s)
			% total CPU time consumed: 10h 19s (36019507965 usecs) (core saturation: 12%)
			& 3,1 & \ETAdag{qqq} & \ETAdag{qqq} \\
			\\
			\\
			\\
			\\
			\\
		\end{tabular}
		\\
		\\
		(a) Verifying HTM locks alone.
		&
		(b) Verifying the lock's client code.
		\end{tabular}
	\end{center}
	\caption{Continuation of Table~\ref{tab:tm-verifs},
		demonstrating abstraction reduction \cite{dbug-phdthesis}
		on the {\tt avl\_fixed} and {\tt map\_basic} tests
		by verifying HTM mutex implementations separately.
		Tested with STM semantics, as both lock implementations include a retry loop.}
	\label{tab:tm-verifs2}
\end{table}

Note two curiosities:
firstly,
the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
(changing a read+write to a read only)
had on even the correctness tests:
{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}),
on account of DPOR no longer needing to reorder the (now) read-read access pairs.
Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
%(CPU time divided by state space size)
than {\tt mutex}
% TODO recalc
(roughly 9 interleavings per second for the former, 22 for the latter),
because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
behind an API Landslide understands,
the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
(\sect{\ref{sec:landslide-blocking}}).
In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.

>>>>

Each preemption point ({\tt struct hax} as defined in {\tt tree.h})
includes three new fields.
The boolean {\tt h->xbegin} is set if the preemption point occurred at an {\tt \_xbegin()} call.
Then if set (as the tag of an option type),
the twin lists of integers
{\tt h->xabort\_codes\_ever}
and
{\tt h->xabort\_codes\_todo}
store possible error codes this {\tt \_xbegin()} call should be tested to possibly return.
The former list stores all such error codes, whether already tested or yet to be tested,
while the latter serves as a workqueue that indicates only those not already tested yet
(serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
The state space estimators (\sect{\ref{sec:landslide-estimate}})
check the length of these lists, in addition to DPOR's tagged threads,
when computing the number of marked children.

@@

To address this,
I extended Landslide
with {\em retry set reduction} (commit 86657c7),
an experimental feature named after prior work's analogous {\em sleep sets}
(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
if either or both sides of a memory conflict were part of a transaction,
it records a {\em retry set},
i.e., the pair of {\tt xbegin} results executed by the conflicting threads,
to accompany that branch.
Like the {\tt xabort\_codes} lists, the state space estimators (\sect{\ref{sec:landslide-estimate}})
check the number of retry sets when counting the number of marked children.
% ah-- not compatible after all!
%in addition to tracking the set of abort codes possible at each preemption point.
Later, when traversing the new branch,
Landslide remembers those threads' expected {\tt xbegin} results
and refuses to test any others
(unless DPOR separately found them, too, to conflict),
thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
%The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
Upcoming in the evaluation, Figure~\ref{fig:retry-reduction} will visualize the reduction achieved in two test cases.

@@

			% note the somewhat chaotic behaviour between quicksand 2,2 and 2,3; the latter surprisingly FABs faster
			% this is likely because non-drpp SSes were not big enough to defer for longer in 2,2,
			% whereas in 2,3 it went straight to the buggy SS asap

@@

\begin{table}[t]
	\begin{center}
		\footnotesize
		\begin{tabular}{cc}
		\begin{tabular}{cc||r|r}
			& & \multicolumn{2}{c}{\bf STM ({\tt -A -S})} \\
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.41} & \ints{4} \\
			({\tt spinlock})
			& 2,2 & \cpu{198.77} & \ints{1702} \\
			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
			& 3,1 & \cpu{35.28} & \ints{246} \\
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.55} & \ints{4} \\
			({\tt spin\_fixed})
			& 2,2 & \cpu{105.57} & \ints{998} \\
			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 3,1 & \cpu{28.27} & \ints{186} \\
			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \ETAdag{16h 26m} & \ETAdag{432628} \\ % was: 30408, 282084 pre mkrun
			% [JOB 4] progress: 286910/432628 brs (66.317896%), ETA 6h 26m 25s (elapsed 9h 59m 48s)
			% total CPU time consumed: 10h 4s (36004288982 usecs) (core saturation: 12%)
			\hline
			{\tt lock}
			& 2,1 & \cpu{3.44} & \ints{4}		\\
			({\tt mutex})
			& 2,2 & \cpu{16.83} & \ints{180}	\\
			& 2,3 & \cpu{405.21} & \ints{9372}	\\
			& 2,4 & \cpu{qqq} & \ints{qqq}	\\ % was: 15266 sec 384512 pre mkrun
			& 3,1 & \cpu{15.00} & \ints{132}	\\
			%& 3,2 & \cpu{29432.87} & \ints{707086}	\\ % pre mkrun fix (let alone bottom1)
			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \cpu{665.89} & \ints{15064}	\\
			\hline
			{\tt lock\_fast}
			& 2,1 & \cpu{3.25} & \ints{1}	\\
			({\tt spin\_fixed})
			& 9,9 & \cpu{4.61} & \ints{1}	\\
			\hline
			{\tt lock\_fast}
			& 2,1 & \cpu{3.19} & \ints{1}	\\
			({\tt mutex})
			& 9,9 & \cpu{4.62} & \ints{1}	\\
			\end{tabular}
			&
			\begin{tabular}{cc||r|r}
			& & \multicolumn{2}{c}{\bf Non-transactional} \\
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt avl\_mutex}
			& 2,1 & \cpu{3.48} & \ints{7} \\
			& 2,2 & \cpu{6.25} & \ints{85} \\
			& 2,3 & \cpu{24.69} & \ints{561} \\
			& 2,4 & \cpu{217.98} & \ints{4984} \\
			& 3,1 & \cpu{8.26} & \ints{129} \\
			& 3,2 & \cpu{1403.46} & \ints{30653} \\
			& 3,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
			& 4,1 & \cpu{199.96} & \ints{4488} \\
			& 4,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
			\cline{1-4}
			{\tt map\_mutex}
			& 2,1 & \cpu{39.81} & \ints{83} \\
			& 2,2 & \ETAdag{26h 21m} & \ETAdag{1085126} \\
			% [JOB 9] progress: 496530/1085126 brs (45.757812%), ETA 16h 21m 0s (elapsed 9h 59m 38s)
			% total CPU time consumed: 10h 19s (36019507965 usecs) (core saturation: 12%)
			& 3,1 & \ETAdag{173d 17h} & \ETAdag{12572187} \\
			% [JOB 9] progress: 496281/12572187 brs (3.947451%), ETA 173d 6h 38m 55s (elapsed 9h 59m 43s)
			% total CPU time consumed: 10h 20s (36020148608 usecs) (core saturation: 12%)
			\\
			\\
			\\
			\\
			\\
		\end{tabular}
		\\
		\\
		(a) Verifying HTM locks alone.
		&
		(b) Verifying the lock's client code.
		\end{tabular}
	\end{center}
	\caption{Continuation of Table~\ref{tab:tm-verifs},
		demonstrating abstraction reduction \cite{dbug-phdthesis}
		on the {\tt avl\_fixed} and {\tt map\_basic} tests
		by verifying HTM mutex implementations separately.
		Tested with STM semantics, as both lock implementations include a retry loop.}
	\label{tab:tm-verifs2}
\end{table}

Note two curiosities:
firstly,
the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
(changing a read+write to a read only)
had on even the correctness tests:
{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by nearly half compared to {\tt lock}({\tt spinlock}),
on account of DPOR no longer needing to reorder the (now) read-read access pairs.
Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
%(CPU time divided by state space size)
than {\tt mutex}
% TODO recalc
(roughly 9 interleavings per second for the former, 22 for the latter),
because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
behind an API Landslide understands,
the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
(\sect{\ref{sec:landslide-blocking}}).
In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.
